---
output: github_document
---

<!-- README.md is generated from README.Rmd, so please edit the latter. -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# mooring

<!-- badges: start -->
[![R-CMD-check](https://github.com/dankelley/mooring/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/dankelley/mooring/actions/workflows/R-CMD-check.yaml)

<!-- badges: end -->

The goal of the `mooring` package is to facilitate computations of the extent
of 'knockdown' of oceanographic moorings caused by ocean currents that exert
drag on mooring elements.

In addition to a fairly broad suite of functions, the package also provides
crude interactive tools for exploring simple moorings.

Conventional R help is provided, in addition to several vignettes.

At the moment, `mooring` only handles bottom-anchored moorings in which the top
buoyancy element is below the sub-surface.

## Installation

You can install the development version of `mooring` from
[GitHub](https://github.com/) by typing the following in an R console.
(Uncomment the first line, if the `devtools` package is not already installed
on your computer.)

``` r
# install.packages("devtools")
devtools::install_github("dankelley/mooring")
```
## Example

The following shows how to assess the knockdown of a mooring consisting of a
bottom anchor, 100 m of wire, and a 20-inch diameter float, in a region of
water depth 120 m with a constant current of 0.5 m/s (roughly 1 knot).  To
learn more, (a) increase the current to 1 m/s and observe changes in the
mooring shape, (b) set the `which` argument of `plot.mooring()`, to focus more
directly on the knockdown, then (c) explore the effects of specifying
depth-dependent currents in the `knockdown()` call.

```{r example}
library(mooring)
# Design a mooring with a 20-inch float attached to a bottom anchor with 100m of wire cable.
m <- mooring(anchor(depth=120), wire(length=100), float("HMB 20"))
# Discretise wire portion (to 1m spacing), then apply a 0.5 m/s current.
md <- discretise(m)
mdk <- knockdown(md, u=0.5)
par(mfrow=c(1, 2))
plot(md, fancy=TRUE)
plot(mdk, fancy=TRUE)
```

