---
output: github_document
bibliography: vignettes/mooring.bib
---

<!-- README.md is generated from README.Rmd, so please edit the latter. -->

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.path = "man/figures/README-",
    out.width = "100%"
)
```

# mooring

<!-- badges: start -->
[![R-CMD-check](https://github.com/dankelley/mooring/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/dankelley/mooring/actions/workflows/R-CMD-check.yaml)

<!-- badges: end -->

The goal of the `mooring` package is to facilitate computations of the extent
of 'knockdown' of oceanographic moorings caused by ocean currents that exert
drag on mooring elements.

In addition to a fairly broad suite of functions, the package also provides
crude interactive tools for exploring simple moorings.

Conventional R help is provided, in addition to several vignettes.

At the moment, `mooring` only handles bottom-anchored moorings in which the top
buoyancy element is below the sub-surface.

## Installation

You can install the development version of `mooring` from
[GitHub](https://github.com/) by typing the following in an R console.
(Uncomment the first line, if the `devtools` package is not already installed
on your computer.)

``` r
# install.packages("devtools")
devtools::install_github("dankelley/mooring")
```
## Example

The following shows how to assess the knockdown of a mooring consisting of a
bottom anchor, 100 m of wire, and a 16-inch Viny float, in a region of
water depth 120 m with a constant current of 1 m/s (roughly 2 knot).  To
learn more, (a) alter the current and observe changes in the
mooring shape, (b) set the `which` argument of `plot.mooring()`, to focus more
directly on the knockdown, then (c) explore the effects of specifying
depth-dependent currents in the `knockdown()` call.

```{r example}
library(mooring)
# Design a mooring with a 16-inch Viny float attached to a bottom
# anchor with 100m of wire cable.
m <- mooring(anchor(depth = 120), wire(length = 100), float("16in Viny"))
# Discretise wire portion (to 1m spacing).
md <- discretise(m)
# Apply a 1 m/s current.
mdk <- knockdown(md, u = 1)
plot(mdk, fancy = TRUE)
```

# Acknowledgements 

The mooring-element properties that are stored in the `mooringElements`
database stem from three main sources. The first was Richard Dewey's excellent
mooring-analysis Matlab software, described by @dewey_mooring_1999. The second
was in spreadsheets and other documents provided by Jay Barthelotte and Matthew
Lawson of the Bedford Institute of Oceanography
([BIO](https://www.bio.gc.ca/index-en.php)). And the third was in spreadsheets
and other documents provided by Danielle Dempsey and Nicole Torrie of the
Centre for Marine Research ([CMAR](https://cmar.ca)). These sources are
designated `"Dewey"`, `"BIO"` and `"CMAR"`, respectively, in `mooringElements`.

Conversations with Clark Richards and Chantelle Layton (both of
[BIO](https://www.bio.gc.ca/index-en.php)) were helpful in identifying data
sources and in clarifying the nature of mooring systems. Documents by
@hamilton_validation_1989 and @hamilton_mooring_1997 cast further light on
important practicalities.

# References
