---
title: Mooring Response to Unidirectional Flow
author: Dan Kelley
date: "`r Sys.Date()`"
bibliography: model.bib
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
    fig_caption: yes
    fig_width: 3.5
    fig_height: 3.5
    dpi: 72
    dev.args: list(pointsize=11)
vignette: >
  %\VignetteIndexEntry{Mooring Response to Unidirectional Flow}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---
\newcommand{\D}{{\cal D}}
\newcommand{\B}{{\cal B}}
\newcommand{\T}{{\cal T}}
\newcommand{\Eq}{Equation}

# Introduction

Anchored oceanographic mooring lines are deformed by currents, in much the same
way as trees are bent by wind.  For very simple moorings, e.g. consisting of a
single float connected by a uniform cable to an anchor at the bottom, the
deformation can be calculated if the current is uniform.  The physics is
analogous to the sagging of telephone lines between poles, and falls into the
general category of a catenary.  However, things are more complicated for
depth-varying currents, and for mooring lines that consist of many elements
connected by cables.

An approach to analysing realistic oceanographic moorings is to use lumped-mass
theory, in which elements such as floats and instrument packages are
parameterized by their buoyancy and drag characteristics, and in which cables
envisioned as connected line elements.  This approach was taken by
@dewey_mooring_1999 in a GUI-based Matlab system.  The present treatment is of
similar foundation, although with some added simplifications (such as
unidirectional current) that are relevant to the basic problem of computing
"knock-down", i.e. the vertical displacement of instruments that is caused
by slanting of the mooring line.

# Model

Consider the case of unidirectional horizontal current aligned with the $x$
axis. Following oceanographic convention, let $z$ be the vertical coordinate,
positive upwards, with $z=0$ at the surface.  The mooring is modelled as a
series of $n$ connected elements, with the top element being numbered $i$.
Water density and velocity are assumed to be a functions of depth, with values
$\rho_i$ and $u_i$ at the location $z_i$ of the $i$-th element.

Each element is described by the following properties.

* Length, $L_i$ [m].  For a portion of cable, this has the common meaning.  For
  a float, instrument, or other thing, it is the contribution that the item
  makes to the overall mooring length.
* Area, $A_i$ [m$^2$] projected in the $x$ direction.
* A measure of buoyancy force divided by $g$, the acceleration due to gravity.
  For a solid object, this is a simple number, expressed in kg, and denoted
  $m_i$ below. For a cable, it is expressed per unit length of cable, i.e. in
  kg/m, and denoted $\mu_i$ below.  Manufacturers tend to provide $m$ or $\mu$
  values in technical documentation about mooring components. Note that they
  refer to in-water buoyancy force, but the distinction between freshwater
  and seawater is seldom made clear.
* Drag coefficient, $C_i$, defined as the ratio of drag force, $\D_i$, to
  $\frac{1}{2}A_i\rho_i u_i|u_i|$.

A tension force exists between each interior mooring element.  The tension
above the $i$-th interior element (thus pulling upward on it) is denoted
$\T_i$, and $\T_{i+1}$ represents the tension pulling down from below.  These
tensions will be directed vertically if there is no current, but they will have
horizontal components with a current, because it will distort the mooring from
its upright orientation.  The angle made by the $i$-th element to the vertical
is represented with $\phi_i$.  Determining these two quantities, $\T$ and
$\phi$, is the key to inferring the response of a mooring to currents.

Quasi-steady dynamics are assumed, so that forces must balance in both the
horizontal and vertical directions. The tension force has components in both
the $x$ and $z$ directions, but the buoyancy force (defined as positive for
objects that would rise if unattached to a mooring) acts only in the vertical,
upwards if $m$ or $\mu$ is positive.

The buoyancy force for a solid object such as an anchor or a float is found
with (Equation 1)
\begin{equation}
\B_i = -g m_i
\end{equation}
while it is represented with (Equation 2)
\begin{equation}
\B_i = -g \mu_i \Delta l_i
\end{equation}
for a portion of a cable of length $\Delta l_i$.

The drag force is expressed with (Equation 3)
\begin{equation}
\D_i = \frac{1}{2} A_i C_i \rho_i u_i |u_i|
\end{equation}
where $A_i$ is the area projected in the flow direction.

With these definitions, a steady-state assumption dictates that forces must be
balanced in the horizontal and vertical directions for each element of the
mooring.

Consider the case of a mooring consisting of $n$ elements attached to an
immovable bottom anchor, and let the index $i=1$ refer to the top element, with
$i$ increasing to $i=n$ at the anchor.

Three forces are involved in this situation: (a) a horizontal drag force, $\D$,
associated with the current, (b) a vertical buoyancy force, $\B$, caused by a
density mismatch between a component and the background water, and (c) a
tension force, $\T$, along the connection between elements.  The tension force
has only a vertical component for an upright mooring, but horizontal components
occur for a mooring that is tilted by a current.

By definition, the top element has no tension from above, and so its force
balance is (Equation 4)
\begin{equation}
\begin{split}
\D_1 &= \T_2 \sin\phi_2\cr
\B_1 &= \T_2 \cos\phi_2
\end{split}
\end{equation}
where $\T_2$ is the tension between this element and the one below it, and
$\phi_2$ is the angle that tension force makes to the vertical.  These two
equations may be solved for $\T_2$ and $\phi_2$, yielding (Equation 5)
\begin{equation}
\begin{split}
\T_2 &= \sqrt{\D_1^2 + \B_1^2}\cr
\phi_2 &= \tan^{-1}\frac{\D_1}{\B_1}
\end{split}
\end{equation}

If there are more than 2 elements in the mooring, this same reasoning, extended
to account for stress and angle below the first element, yields (Equation 6)
\begin{equation}
\begin{split}
\D_2 + \T_2 \sin\phi_2 &= \T_3 \sin\phi_3\cr
\B_2 + \T_2 \cos\phi_2 &= \T_3 \cos\phi_3
\end{split}
\end{equation}
and, since $\phi_2$ and $\T_2$ are known from Equation 5, Equation 6 may be solved
with (Equation 7)
\begin{equation}
\begin{split}
\T_3 &= \sqrt{(\D_2+\T_2\sin\phi_2)^2 + (\B_2+\T_2\cos\phi_2)^2}\cr
\phi_3 &= \tan^{-1}\frac{\D_2+\T_2\sin\phi_2}{\B_2+\T_2\cos\phi_2}
\end{split}
\end{equation}

This may be generalized to larger moorings, with (Equation 8)
\begin{equation}
\begin{split}
\T_i &= \sqrt{(\D_{i-1}+\T_{i-1}\sin\phi_{i-1})^2+(\B_{i-1}+\T_{i-1}\cos\phi_{i-1})^2}\cr
\phi_i &= \tan^{-1}\frac{\D_{i-1}+\T_{i-1}\sin\phi_{i-1}}{\B_{i-1}+\T_{i-1}\cos\phi_{i-1}}
\end{split}
\end{equation}
applying at each element, $i$.

This leads to a simple way to describe mooring response to a depth-independent
current, in which $D_i$ and $B_i$ are known quantities for each $i$ from 1 to
$n$:
1. First, use Equation 5 to compute $\T_2$ and $\phi_2$, the angle and tension below
   the top element.
2. Then, use Equation 8 to compute $\T_i$ and $\phi_i$ for $i=3$, then $i=4$, etc.
   until reaching $i=n$.

(Note that $\T_1$ and $\phi_1$ are not defined in this notation.)

If the current varies with depth, then the drag of a given element will depend
on depth of the element, which in turn depends on the shape of the mooring.
This may be handled by an iterative scheme.  In the first iteration, the
vertical positions of mooring elements are inferred by summing their heights,
and the above procedure is used to infer $\phi_i$. Then, these angles are used
to calculate a second estimate of element locations, and the procedure is
repeated using drags appropriate to those locations.  This is repeated until
the solution converges to within a stated criterion.

# Example

```{r echo=FALSE}
library(mooring)
f <- float("HMB 20")
w <- wire(length=100)
```


Figure 1 shows the result of simulating a 20-inch Hydro Float Mooring Buoy
(with $`r f[[1]]$buoyancy`$kg buoyancy) connected to a bottom anchor with 100 m
of quarter-inch jacketed wire (with $`r w[[1]]$buoyancy`$kg/m of buoyancy), in
a depth-independent 0.5 m/s (roughly 1 knot) current.  Note that the float is
predicted to sink $7.7$m with this current.  However, the predicted knockdown
increases to $41.2$m with a $1$m/s current, suggesting that this float would be
of limited utility in supporting an oceanographic mooring a region with strong
currents.

```{r echo=FALSE,dev.args=list(pointsize=9),fig.cap="Response of a mooring to a $u=1$m/s current, as described in the text. **Left**: Cable tension, in kg, with gray for the $u=0$ case and black for the $u=1$m/s case. **Right:** Mooring shape, with same colour scheme as left panel."}
source("example_01.R")
```


# Special case: negligible currents

As a special case, if $u_i=0$ for all $i$, then there will be no drag, and so
(assuming adequate buoyancy), the mooring will be aligned in the vertical, and
the vertical equation becomes
\begin{equation}
\T_{i+1} = \T_i + \B_i
\end{equation}
which yields
\begin{equation}
\T_i = \sum_{1}^{i-1} \B_j
\end{equation}
for $i>1$.  Physically, this means that the downward tension at any level
balances the total upward buoyancy force of all the elements above it. The
bottom tension, $\T_n$, may be interpreted as the minimum anchor weight that
will hold the mooring in place in the absence of currents. However, it should
be obvious that heavier weights will be required for practical situations.

# References


