---
title: Default Values for Mooring Elements
author: "Dan Kelley (https://orcid.org/0000-0001-7808-5911)"
date: "`r Sys.Date()`"
bibliography: mooring.bib
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
    fig_caption: yes
    fig_width: 3.5
    fig_height: 3.5
    dpi: 72
    dev.args: list(pointsize=11)
vignette: >
  %\VignetteIndexEntry{Default Values for Mooring Elements}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The data stored in `data(mooringElements)`, key features of which are presented
below, are based on the following sources.  Each entry has a column named
`name` that indicates how the item is designated in calls to functions within
the package, and another column named `OriginalName` that indicates the name
used in the original source.

1. A Matlab file named `mdcodes.mat`, contained within Dewey's [Matlab-based
   mooring package](http://web.uvic.ca/~rdewey/mooring/moordyn.php) (which was
   the inspiration for the present package). These entries have "Dewey" in the
   source column in the tables. The details of data provenance were not
   pursued, in creating the present R package.
2. Information obtained from [Bedford Institute of
   Oceanography](https://www.bio.gc.ca/index-en.php) ("BIO" in the source
   column), based in Dartmouth, Nova Scotia, Canada.
3. Information obtained from the [Centre for Marine Applied
   Research](https://cmar.ca) ("CMAR" in the tables), based in Dartmouth, Nova
   Scotia, Canada.

## Anchor characteristics

The `buoyancy` values are in kg and the `height` values are in m.

```{r echo=FALSE}
data(mooringElements, package = "mooring")
for (name in names(mooringElements)) {
    mooringElements[[name]]$code <- NULL
}
```
```{r echo=FALSE}
o <- order(mooringElements$anchors$name)
knitr::kable(mooringElements$anchors[o, ], row.names = FALSE)
```

## Chain characteristics

The `buoyancyPerMeter` values are in kg/m, while `areaPerMeter` is in m, and
`CD` is unitless.  For the Dewey data, area is computed from stated chain
width.

```{r echo=FALSE}
o <- order(mooringElements$chains$name)
knitr::kable(mooringElements$chains[o, ], row.names = FALSE)
```


## Connector characteristics

The `buoyancy` values are in kg, while `height` and `width` are in m, and `CD`
is unitless.  Not all data sources list `height` and `width` (or comparable
values) and so zeros are used in such instances.  Since the only use of these
values in the package are to compute area, for a drag calculation, and since
the drag on floats and wires is likely to exceed that on small elements, this
zeroing-out may not pose a large problem.

```{r echo=FALSE}
o <- order(mooringElements$connectors$name)
knitr::kable(mooringElements$connectors[o, ], row.names = FALSE)
```


## Float characteristics

The `buoyancy` values are in kg, while `height` and `diameter` are in m, and
`CD` is unitless.  For the Dewey data, area is computed from stated float
diameter.

```{r echo=FALSE}
o <- order(mooringElements$floats$name)
knitr::kable(mooringElements$floats[o, ], row.names = FALSE)
```


## Instrument characteristics

The `buoyancy` values are in kg, while `area` is in \eqn{m^2}{m^2}, and `CD` is
unitless.

```{r echo=FALSE}
o <- order(mooringElements$instruments$name)
knitr::kable(mooringElements$instruments[o, ], row.names = FALSE)
```

## Misc characteristics

The `buoyancy` values are in kg, while `height` is in m, `area` is in m^2, and
`CD` is unitless.

```{r echo=FALSE}
o <- order(mooringElements$miscs$name)
knitr::kable(mooringElements$miscs[o, ], row.names = FALSE)
```


## Release characteristics

The `buoyancy` values are in kg, while `height` and `width` are in m, and `CD`
is unitless.

```{r echo=FALSE}
o <- order(mooringElements$releases$name)
knitr::kable(mooringElements$releases[o, ], row.names = FALSE)
```

## Wire characteristics

The `buoyancyPerMeter` values are in kg/m, while `areaPerMeter` is in m, and
`CD` is unitless.  For the Dewey data, area is computed from stated wire width.

```{r echo=FALSE}
o <- order(mooringElements$wires$name)
knitr::kable(mooringElements$wires[o, ], row.names = FALSE)
```
