---
title: Default Values for Mooring Elements
author: Dan Kelley
date: "`r Sys.Date()`"
bibliography: model.bib
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
    fig_caption: yes
    fig_width: 3.5
    fig_height: 3.5
    dpi: 72
    dev.args: list(pointsize=11)
vignette: >
  %\VignetteIndexEntry{Default Values for Mooring Elements}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

The data stored in `data(mooringElements)` were extracted from Dewey's Matlab
matlab file named `mdcodes.mat`, which were apparently extracted from more
primary sources. Neither the provenance and accuracy of individual items was
checked before inclusion here.

Dewey reports dimensions in centimetres, but these are converted to metres
here.

The column names are not the same as Dewey uses.  The meaning of the final
column is unknown, so it is just called `code` here.

## Float characteristics

The `buoyancy` values are in kg, the next three quantities are in m, and `CD`
is unitless.

Dewey lists meaningless values in the column named `width` here, and those
values are not used in this code provided in this package, although they are
retained in the dataset.

```{r echo=FALSE}
data(mooringElements, package="mooring")
knitr::kable(mooringElements$floats)
```

## Wire characteristics

The `buoyancy` values are in kg/m, the next three quantities are in m, and `CD`
is unitless.

Dewey lists meaningless values in the columns that are named `height` and
`diameter` here, and those values are not used in this code provided in this
package, although they are retained in the dataset.

```{r echo=FALSE}
knitr::kable(mooringElements$wires)
```

## Chain characteristics


The `buoyancy` values are in kg, the next three quantities are in m, and `CD`
is unitless.

Dewey lists meaningless values in the column that is named `diameter` here, and
those values are not used in this code provided in this package, although they
are retained in the dataset.

```{r echo=FALSE}
knitr::kable(mooringElements$chains)
```

