---
title: Default Values for Mooring Elements
author: Dan Kelley
date: "`r Sys.Date()`"
bibliography: model.bib
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
    fig_caption: yes
    fig_width: 3.5
    fig_height: 3.5
    dpi: 72
    dev.args: list(pointsize=11)
vignette: >
  %\VignetteIndexEntry{Default Values for Mooring Elements}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

The data stored in `data(mooringElements)` were extracted from Dewey's Matlab
matlab file named `mdcodes.mat`, which were apparently extracted from more
primary sources. Neither the provenance and accuracy of individual items was
checked before inclusion here.

Dewey reports dimensions in centimetres, but these are converted to metres
here.

The column names are not the same as Dewey uses.  The meaning of the final
column is unknown, so it is just called `code` here.

## Anchor characteristics

The `buoyancy` values are in kg and the `height` values are in m.

```{r echo=FALSE}
data(mooringElements, package="mooring")
o <- order(mooringElements$anchors$name)
knitr::kable(mooringElements$anchors[o,])
```

## Chain characteristics

The `buoyancy` values are in kg, while `height` and `width` are in m, and `CD`
is unitless.  The `code` values are listed for data from Dewey only, and they
are not used in this package.

```{r echo=FALSE}
data(mooringElements, package="mooring")
o <- order(mooringElements$chains$name)
knitr::kable(mooringElements$chains[o,])
```


## Float characteristics

The `buoyancy` values are in kg, the next three quantities are in m, and `CD`
is unitless.

Dewey lists meaningless values in the column named `width` here, and those
values are not used in this code provided in this package, although they are
retained in the dataset.

```{r echo=FALSE}
data(mooringElements, package="mooring")
o <- order(mooringElements$floats$name)
knitr::kable(mooringElements$floats[o,])
```

## Release characteristics

The `buoyancy` values are in kg, while `height` and `width` are in m, and `CD`
is unitless.  The `code` values are unused in this package.

```{r echo=FALSE}
data(mooringElements, package="mooring")
o <- order(mooringElements$releases$name)
knitr::kable(mooringElements$releases[o,])
```

## Wire characteristics

The `buoyancyPerMeter` values are in kg/m, while `diameter` is in m, and `CD`
is unitless.  Note that Dewey sets diameter to 0 for wires, saves diameter in a
width field, and lists a nonsense value for height, evidently so his code can
distinguish floats from wires; we needn't copy this convention the data are
stored in `list` objects.

```{r echo=FALSE}
data(mooringElements, package="mooring")
o <- order(mooringElements$wires$name)
knitr::kable(mooringElements$wires[o,])
```
