---
title: Default Values for Mooring Elements
author: "Dan Kelley (https://orcid.org/0000-0001-7808-5911)"
date: "`r Sys.Date()`"
bibliography: mooring.bib
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
    fig_caption: yes
    fig_width: 3.5
    fig_height: 3.5
    dpi: 72
    dev.args: list(pointsize=11)
vignette: >
  %\VignetteIndexEntry{Default Values for Mooring Elements}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The data stored in `data(mooringElements)` were extracted from Dewey's Matlab
matlab file named `mdcodes.mat`, which were apparently extracted from more
primary sources. Neither the provenance and accuracy of individual items was
checked before inclusion here.

Dewey reports dimensions in centimetres, but these are converted to metres
here.

The column names are not the same as Dewey uses.  The meaning of the final
column is unknown, so it is just called `code` here, and its values are not are
used in this package.

## Anchor characteristics

The `buoyancy` values are in kg and the `height` values are in m.  Note that
the values shown below have no affect on the computation of mooring knockdown.

```{r echo=FALSE}
data(mooringElements, package="mooring")
o <- order(mooringElements$anchors$name)
knitr::kable(mooringElements$anchors[o,], row.names=FALSE)
```

## Chain characteristics

The `buoyancyPerMeter` values are in kg/m, while `height` and `width` are in m,
and `CD` is unitless.  The `code` values are listed for data from Dewey only,
and they are not used in this package.

```{r echo=FALSE}
data(mooringElements, package="mooring")
o <- order(mooringElements$chains$name)
knitr::kable(mooringElements$chains[o,], row.names=FALSE)
```


## Connector characteristics

The `buoyancy` values are in kg, while `height` and `width` are in m, and `CD`
is unitless.  Not all data sources list `height` and `width` (or comparable values) and so zeros are used in such instances.  Since the only use of these values in the package are to compute area, for a drag calculation, and since the drag on floats and wires is likely to exceed that on small elements, this zeroing-out may not pose a large problem. The `code` values are listed for data from Dewey only, and they
are not used in this package.

```{r echo=FALSE}
data(mooringElements, package="mooring")
o <- order(mooringElements$connectors$name)
knitr::kable(mooringElements$connectors[o,], row.names=FALSE)
```


## Float characteristics

The `buoyancy` values are in kg, while `height` and `diameter` are in m, and
`CD` is unitless.  The `code` values are listed for data from Dewey only, and
they are not used in this package.

```{r echo=FALSE}
data(mooringElements, package="mooring")
o <- order(mooringElements$floats$name)
knitr::kable(mooringElements$floats[o,], row.names=FALSE)
```


## Instrument characteristics

The `buoyancy` values are in kg, while `area` is in \eqn{m^2}{m^2}, and `CD` is
unitless.  The `code` values are listed for data from Dewey only, and they are
not used in this package.

```{r echo=FALSE}
data(mooringElements, package="mooring")
o <- order(mooringElements$instruments$name)
knitr::kable(mooringElements$instruments[o,], row.names=FALSE)
```


## Release characteristics

The `buoyancy` values are in kg, while `height` and `width` are in m, and `CD`
is unitless.  The `code` values are listed for data from Dewey only, and they
are not used in this package.

```{r echo=FALSE}
data(mooringElements, package="mooring")
o <- order(mooringElements$releases$name)
knitr::kable(mooringElements$releases[o,], row.names=FALSE)
```

## Wire characteristics

The `buoyancyPerMeter` values are in kg/m, while `diameter` is in m, and `CD`
is unitless.  The `code` values are listed for data from Dewey only, and they
are not used in this package.

Note that Dewey sets diameter to 0 for wires, saves diameter in a width field,
and lists a nonsense value for height.  These are signals that let his code
distinguish between floats, wires, etc., but the present package handles this
differently.

```{r echo=FALSE}
data(mooringElements, package="mooring")
o <- order(mooringElements$wires$name)
knitr::kable(mooringElements$wires[o,], row.names=FALSE)
```
