---
title: Default Values for Mooring Elements
author: Dan Kelley
date: "`r Sys.Date()`"
bibliography: model.bib
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
    fig_caption: yes
    fig_width: 3.5
    fig_height: 3.5
    dpi: 72
    dev.args: list(pointsize=11)
vignette: >
  %\VignetteIndexEntry{Default Values for Mooring Elements}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

The data stored in `data(mooringElements)` were extracted from Dewey's Matlab
matlab file named `mdcodes.mat`, which were apparently extracted from more
primary sources. Neither the provenance and accuracy of individual items was
checked before inclusion here. Note that Dewey reports dimensions in
centimetres, but these are converted to metres here.

## Float characteristics

The column names are not the same as in Dewey's Matlab file, although the
meaning of the final column is unknown, so it is just called `code` here.

The `buoyancy` values are in kg, the next three quantities are in m, and `CD`
is unitless.

```{r echo=FALSE}
data(mooringElements, package="mooring")
knitr::kable(mooringElements$floats)
```

## Wire characteristics

The column names are not the same as in Dewey's Matlab file, although the
meaning of the final column is unknown, so it is just called `code` here.

The `buoyancy` values are in kg/m, the next three quantities are in m, and `CD`
is unitless.  The `width` column indicates the wire width in m.

Note that Dewey lists meaningless values in the columns that are named `height`
and `diameter` here, and those values are not used in this code provided in
this package, although they are retained in the dataset.

```{r echo=FALSE}
knitr::kable(mooringElements$wires)
```

## Cable characteristics


The column names are not the same as in Dewey's Matlab file, although the
meaning of the final column is unknown, so it is just called `code` here.

The `buoyancy` values are in kg/m, the next three quantities are in m, and
`CD` is unitless.

Note Dewey lists meaningless values in the column that is named `diameter`
here, and those values are not used in this code provided in this package,
although they are retained in the dataset.

```{r echo=FALSE}
knitr::kable(mooringElements$chains)
```

