---
title: Object-orientation Scheme
author: "Dan Kelley (https://orcid.org/0000-0001-7808-5911)"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
    fig_caption: yes
    fig_width: 3.5
    fig_height: 3.5
    dpi: 72
    dev.args: list(pointsize=11)
vignette: >
  %\VignetteIndexEntry{Object-orientation Scheme}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The S3 scheme is used for objects created by functions in the `mooring`
package. There are two types of functions, and two classes of
return value.

The overall functions, `mooring()`, `discretise()` and `knockdown()` all
produce return lists that have class `"mooring"`. The elements of those lists
are produced with what are called "elementary" functions, which include
`anchor()`, `wire()`, `float()`, etc., all of which return objects with main
class `"mooringElement"` and sub-class patterned on the function name.  

The return values from the elementary functions vary with type, and allowances
for these different aspects are made in `knockdown()` and the functions
it accesses, such as `area()`, etc.

For example,
```{r}
library(mooring)
m <- mooring(anchor(), wire(length=75), float(), waterDepth = 100)
```
produces a list with 3 elements, one that holds information on the anchor, a
second that holds information on the wire, and a third that holds information
on the float.  

To get a quick overview of a function, use e.g.
```{r}
m
```
and, for more details, try using `str(m)`.


