---
title: Object-orientation Scheme
author: "Dan Kelley (https://orcid.org/0000-0001-7808-5911)"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
    fig_caption: yes
    fig_width: 3.5
    fig_height: 3.5
    dpi: 72
    dev.args: list(pointsize=11)
vignette: >
  %\VignetteIndexEntry{Object-orientation Scheme}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The [S7](https://cran.r-project.org/web/packages/S7/index.html) scheme
is used for objects created by functions in the `mooring` package.

The functions `mooring()`, `segmentize()` and `knockdown()` all deal
with moorings, and they return an object that is named so that other
functions will know how to deal with the information.  Moorings
consist of what are called "elements" in the package, and these are
created with a suite of specialized functions.  In common work, the
main element-creating functions are `anchor()` for creating anchors,
`wire()` for creating wires, `instrument()` for creating instruments,
and `float()` for creating floats.  Consult the "see also" section of
the documentation for any of these functions, to discover other
element-creating functions.

Some examples may clarify.  First, we create an anchor element,
and find its contents.

```{r}
library(mooring)
a <- anchor()
```

To see it's contents, simply type its name in an interactive console:
```{r}
a
```

The individual components can be accessed with the `@` syntax used by
[S7](https://cran.r-project.org/web/packages/S7/index.html). For
example, the buoyancy (the negative of the 'weight' in water, 
expressed in kg) as

```{r}
a@buoyancy
```

You may also alter the components easily; e.g. the following
reduces the height of this anchor to zero.

```{r}
a@height <- 0
a
```

Using e.g. `anchor("?")` gives a list of built-in anchor types,
and it is not uncommon to construct mooring elements by staring
with a known element and changing some quantities.

Turning to moorings, a 4-element case will suffice for
illustration.

```{r}
m <- mooring(anchor(), wire(length = 75), float(), waterDepth = 100)
```

As the reader has no doubt guessed, the way to find the contents is
to name the object.

```{r}
m
```

The way to get the top element is e.g.
```{r}
m@elements[[1]]
```
