---
title: Test Case with Table Tennis Ball
author: "Dan Kelley (https://orcid.org/0000-0001-7808-5911)"
date: "`r Sys.Date()`"
bibliography: mooring.bib
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
    fig_caption: yes
    fig_width: 7
    fig_height: 4
    dpi: 72
    dev.args: list(pointsize=11)
vignette: >
  %\VignetteIndexEntry{Test Case with Table Tennis Ball}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

**Abstract.** This vignette outlines a test of the model against
theory for a very simple mooring, and ends with an idea for an
instrument to measure flow based on mooring dynamics.

# Introduction

As a test of the `mooring` code, and a possible motivation for a
simple experiment, consider a table-tennis ball attached with a thin
fishing line to a solid immovable object. These balls are very buoyant
in water, consisting of a light and thin material with air inside. By
contrast, fishing line is nearly neutrally buoyant in water.  The area
of the ball will appreciably exceed that of the line, unless the line
is very long, and therefore the drag on the ball will be much larger
than the drag on the line.

# Theory

For simplicity, we may assume that the line has zero drag, and zero buoyancy.

Let $L$ be the length of the line and $R$ the radius of the ball. Let
the water density be $\rho$.  Assuming that the density of the ball
(and air inside) is much less than water density, we may approximate
the ball buoyancy force as

\begin{equation}
\tag{1}
B = \frac{4}{3}\pi \rho g R^3
\end{equation}

where $g$ is the acceleration due to gravity.  The drag on the ball
may be approximated as

\begin{equation}
\tag{2}
D = \frac{1}{2} C_D \rho \pi R^2 u^2
\end{equation}

where $C_D$ is a drag coefficient and $u$ is the water speed in the
$x$ direction.

We assume a steady state. If $\tau$ represents tension in the fishing
line, and $\phi$ the angle the line makes to the vertical, then force
balance in the horizontal and vertical directions requires that

\begin{equation}
\tag{3}
\tau\sin\phi = D
\end{equation}

and

\begin{equation}
\tag{4}
\tau\cos\phi = B
\end{equation}
respectively.  Dividing these yields
\begin{equation}
\tag{5}
\phi = \tan^{-1}\frac{D}{B}
\end{equation}

From the above, it follows that tension in the line is

\begin{equation}
\tag{6}
\tau = \frac{B}{\cos\phi}
\end{equation}

that the ball is at position

\begin{equation}
\tag{7}
x' = L\sin\phi
\end{equation}

and that it is at height

\begin{equation}
\tag{8}
z' = L - L\cos\phi
\end{equation}

above the bottom.

It is easy to imagine an experimental setup in which $x'$ might be
measured (replace the anchor with a rod held vertical in the water
column, with the line attached at the bottom, and sight downwards to
get the distance of the float from the rod).

# Test

The predictions can be tested against the `knockdown()` function as
follows. This imagines a river of depth 0.5m (about knee-deep) and a
1m/s current (about walking speed).  Note that the prediction of the
`mooring` code matches that (circled in blue) of the formulae given
above.

```{r fig.cap="*Test of package against direct computation. The black dot is the float position computed in the package, whereas the red circle is the result of direct computation using equations 7 and 8.*"}
library(testthat)
L <- 0.5 # tank depth, m
diameter <- 0.04 # ping-pong balls have 40mm diameter
R <- diameter / 2
area <- pi * R^2
volume <- 4 / 3 * pi * R^3
rho <- 1027
deltaRho <- rho # water vs air (round numbers)
g <- 9.81
B <- g * deltaRho * volume # buoyancy force [N]
CD <- 1
u <- 0.5 # speed, m/s
D <- 0.5 * CD * area * rho * u^2 # ddrag force [N]
phi <- atan2(D, B)
x <- L * sin(phi)
l <- L * cos(phi)
tau <- B / cos(phi)

library(mooring)
f <- float("pingpong", B / g, height = 0, area = pi * R^2, CD = 1)
expect_equal(B / g, buoyancy(f))
expect_equal(D, drag(f, u = u))

w <- wire("gossamer", length = L, buoyancyPerMeter = 0, areaPerMeter = 0, CD = 0)
a <- anchor("fake", buoyancy = -1000, height = 0, CD = 0)
m <- mooring(a, w, f, waterDepth = L)

ms <- segmentize(m, L / 10)
msk <- knockdown(ms, u = u)
draw(msk, fancy = TRUE)
# Check against theory
points(x, L - l, col = 2, cex = 2)
```

# Results

```{r,echo=FALSE}
xcm <- 100 * head(x(msk), 1)
zcm <- -100 * head(z(msk), 1)
```

The theory and the calculations using `knockdown()` agree well.  There
are three main observables that might be part of an experiment to test
this in a natural environment.

1. The horizontal shift of the float, $x'$.  Here, a $`r u`$ m/s
   current is predicted to yield $x'=`r sprintf("%.1f",xcm)`$ cm.
2. The vertical float displacement over the zero-velocity case, $z'$.
   Here, the prediction is $`r sprintf("%.1f",zcm)`$ cm.
3. The tension in the line, $\tau$, is predicted to be $`r
   round(tau,3)`$ Newtons, so that an anchor would need to have mass
   exceeding `r round(1000*tau/g,1)` g to avoid lifting by the drag.
   This mass rises to 74.3 g in a 1 m/s current, which is slightly
   less than the mass of a 3/4-inch bolt, 136 g, and well under the
   weight of a 1-inch bolt, 195 g.  This suggests that a 1-inch bolt
   might be a good anchor for a demonstration of this effect in a
   flume.  (See <https://www.portlandbolt.com/products/nuts/hex-nuts/>
   for a table of bolt properties.)

# Measuring currents

Any of $x'$, $z'$ and $\tau$ could be measured in a classroom or
laboratory setting, but measuring tension might be harder in a field
test, and so a field apparatus should focus on measuring $x'$ and
$x'$. This would not be difficult, for this range of current, using a
metre-stick, to which is attached a table-tennis ball with 1 m of thin
mono-filament fishing line.  A value of $z'$ could be obtained by
lowering the rod slowly, until the float just sinks below the surface,
and observing water level on the stick's distance markings.  A value
of $x'$ could be obtained with a second meter stick held horizontally
so that the ball position could be observed from overhead.

In a laboratory setting, or with a bit of engineering for a field
setting, the tension could be measured.  An advantage of that method
would be that it could be digitized and, using a microcontroller to do
the calculations shown here, a readout could show the current speed.
