---
title: Test Case with Table Tennis Ball
author: "Dan Kelley (https://orcid.org/0000-0001-7808-5911)"
date: "`r Sys.Date()`"
bibliography: mooring.bib
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
    fig_caption: yes
    fig_width: 7
    fig_height: 4
    dpi: 72
    dev.args: list(pointsize=11)
vignette: >
  %\VignetteIndexEntry{Test Case with Table Tennis Ball}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

**Abstract.** This vignette outlines a test of the model against theory for a
very simple mooring, and ends with an idea for an instrument to measure flow
based on mooring dynamics.

# Introduction

As a test of the `mooring` code, and a possible motivation for a simple
experiment, consider a table-tennis ball attached with a thin fishing line to a
solid immovable object. These balls are very buoyant in water, consisting of a
light and thin material with air inside. By contrast, fishing line is nearly
neutrally buoyant in water.  The area of the ball will appreciably exceed that
of the line, unless the line is very long, and therefore the drag on the ball
will be much larger than the drag on the line.

# Theory

For simplicity, we may assume that the line has zero drag, and zero buoyancy.

Let $L$ be the length of the line and $R$ the radius of the ball. Let the water
density be $\rho$.  Assuming that the density of the ball (and air inside) is
much less than water density, we may approximate the ball buoyancy force as
\begin{equation}
B = \frac{4}{3}\pi \rho g R^3
\end{equation}
where $g$ is the acceleration due to gravity.  The drag on the ball may be
approximated as
\begin{equation}
D = \frac{1}{2} C_D \rho \pi R^2 u^2
\end{equation}
where $C_D$ is a drag coefficient and $u$ is the water speed in the $x$ direction.

We assume a steady state. If $T$ represents tension in the fishing line, and
$\phi$ the angle the line makes to the vertical, then force balance in the
horizontal and vertical directions requires that
\begin{equation}
T\sin\phi = D
\end{equation}
and
\begin{equation}
T\cos\phi = B
\end{equation}
respectively.  Dividing these yields
\begin{equation}
\phi = \tan^{-1}\frac{D}{B}
\end{equation}

From the above, it follows that tension in the line is
\begin{equation}
T = \frac{B}{\cos\phi}
\end{equation}
and that the ball is at position
\begin{equation}
x' = L\sin\phi
\end{equation}
and that it is at height
\begin{equation}
z' = L - L\cos\phi
\end{equation}
above the bottom.

It is easy to imagine an experimental setup in which $x'$ might be measured
(replace the anchor with a rod held vertical in the water column, with the line
attached at the bottom, and sight downwards to get the distance of the float
from the rod).

# Test

The predictions can be tested against the `knockdown()` function as follows.
This imagines a river of depth 0.5m (about knee-deep) and a 1m/s current (about
walking speed).  Note that the prediction of the `mooring` code matches
that (circled in blue) of the formulae given above.

```{r}
library(testthat)
L <- 0.5                               # tank depth, m
diameter <- 0.04 # ping-pong balls have 40mm diameter
R <- diameter / 2
area <- pi*R^2
volume <- 4/3*pi*R^3
rho <- 1027
deltaRho <- rho                        # water vs air (round numbers)
g <- 9.8
B <- g * deltaRho * volume             # Newtons
CD <- 1
u <- 0.5                               # speed, m/s
D <- 0.5 * CD * area * rho * u^2       # Newtons
phi <- atan2(D, B)
x <- L * sin(phi)
l <- L * cos(phi)
T <- B / cos(phi)

library(mooring)
f <- float("pingpong", B/g, height=0, area=pi*R^2, CD=1)
expect_equal(B/g, buoyancy(f))         # convert to kg to compare
expect_equal(D/g, drag(f, u=u))        # convert to kg to compare

w <- wire("gossamer", length=L, buoyancyPerMeter=0, areaPerMeter=0, CD=0)
a <- anchor("fake", buoyancy=-1000, height=0, depth=L, CD=0)
m <- mooring(a, w, f)

n <- 500 # number of sub-segments to use
md <- discretise(m, L/n)
mdk <- knockdown(md, u=u)
plot(mdk, fancy=TRUE)
# Check against theory
points(x, L-l, col=2, cex=2)
```

# Results

```{r,echo=FALSE}
xcm <- 100 * head(x(mdk),1)
zcm <- -100 * head(z(mdk),1)
```

The theory and the calculations using `knockdown()` agree well.  There are
three main observables that might be part of an experiment to test this in a
natural environment.

1. The horizontal shift of the float, $x'$.  Here, a $`r u`$ m/s current is
   predicted to yield $x'=`r sprintf("%.1f",xcm)`$ cm.
2. The vertical float displacement over the zero-velocity case, $z'$. Here, the
   prediction is $`r sprintf("%.1f",zcm)`$ cm.
3. The tension in the line, $T$, is predicted to be $`r round(T,3)`$ Newtons,
   so that an anchor would need to have mass exceeding `r round(1000*T/g,1)` g
   to avoid lifting by the drag. This mass rises to 74.3 g in a 1 m/s current,
   which is slightly less than the mass of a 3/4-inch bolt, 136 g, and well
   under the weight of a 1-inch bolt, 195 g.  This suggests that a 1-inch bolt
   might be a good anchor for a demonstration of this effect in a flume.  (See
   <https://www.portlandbolt.com/products/nuts/hex-nuts/> for a table of
   bolt properties.)

# Measuring currents

Any of $x'$, $z'$ and $T$ could be measured in a classroom or laboratory
setting, although it might be harder to measure the tension force in the field,
so a field apparatus should focus on measuring $x'$ and $x'$.  This would not
be difficult, for this range of current, using a metre-stick, to which is
attached a table-tennis ball with 1 m of thin mono-filament fishing line.  A
value of $z'$ could be obtained by lowering the rod slowly, until the float
just sinks below the surface, and observing water level on the stick's distance
markings.  A value of $x'$ could be obtained with a second meter stick held
horizontally so that the ball position could be observed from overhead.

In a laboratory setting, or with a bit of engineering for a field setting, the
tension could be measured.  An advantage of that method would be that it could
be digitized and, using a microcontroller to do the calculations shown here, a
readout could show the current speed.



