---
title: Ingesting object data from a BIO excel file
author: Dan Kelley
date: 2021-05-28
---


There is a .R file here in which I tried to decode things automatically.  This
proved very difficult, because the meanings of things were (I think) indicated
by colour underlays.  Also, there were *lots* of weird things in the file, like
text in numeric fields, negative lengths, etc.  In the end, I decided that
hand-editing would be superior.

I used lower-case, which is easier to type, and fits with the Dewey names.  I
also removed the `"` symbols, which are hard to type in code, and appear in
some instances and not in others.


# wires_bio.csv

There are just two listings.

## Buoyancy/metre

The 3/16 case has -0.76 in a column named "W Nt/m", so I've converted that to
-0.76/9.8=-0.0776 (rounded to 4 places), to get kg/m instead of N/m.

There is no listing for the 1/4 case, but I estimated (guessed) that as

```{r}
round(-0.0776 * ((4/16)/(3/16))^2, 4)
```
This is based on an assumption that the metal inside is similar, etc.

## Diameter

For the diameter, I looked at the column labeled `A (m*2/m)`. I assume that is
the horizontally-projected area. To test that, I computed 
```{r}
16*c(0.00635,0.0079375)/0.0254
```
which yields the expected, for the outside diameters in 16-ths of inch.

## CD

There are no entries in the file and I did not know what `DRAG -N` and `DRAG
-T` meant and so I just used the value $C_D=1.3$, used in Dewey's work.

## Result for wire
```{r}
knitr::kable(read.csv("wires_bio.csv"))
```


# wires_bio.csv

These seem to be in the yellow-background part of the spreadsheet.

The first of these, named `"new glass streamlined float"`, has no listed weight
in pounds or newtons, so I ignored that entry.

There are two length columns in the file, but the values are either identical
or relatable by rounding, so I could choose either.  But is that in the x or
the y direction?

Two columns contain things related to area, one with name `A (m*2/m)` and other
with name `AW (m*2)`.  I think the former is the projected area per meter of z
(as it is for wires).  So, if we assume that the length is in the z direction,
or that the object is roughly spherical, we can infer what the package needs,
namely `height` and `width`, from the ratio of the second area to the first
length.  Whether that's sensible, I just don't know.
```{r}
A<-c(0.164,0.5,0.073,0.5,0.164,1.5,1.5,1.86,2.84,2.474,1.67,2.7,2.7)
L<-c(0.74,0.563,1,0.563,1.187,1,1,1.286,1.286,1.286,1.286,2.42,2.42)
round(A/L,3)
```

```{r}
knitr::kable(read.csv("floats_bio.csv"))
```

