---
title: Ingesting object data from a BIO excel file
author: Dan Kelley
date: 2021-05-28, 2021-06-08, and 2021-06-09
output:
    pdf_document:
        toc: true
        number_sections: true
        extra_dependencies:
            fullpage
---

# Document Revisions

1. 2021-06-08 (thanks to @formulafunracing and @clayton33 for advice on this)
   1. Move discrete chain items like shackles to a new entity, "connector".
   2. Make `chain` objects have `length`, not `height`.
   3. Rename some items for consistency and readability.
   4. Add `originalName` column.

# Introduction

This directory contains a file named `bio.R` in which I tried to decode things
automatically.  This proved to be difficult, because the meanings of things are
(I gather) indicated by colour underlays, and I don't know how to catch that
nuance in a program.  Another problem was that there are *lots* of weird things
in the excel file, like text in numeric fields, negative lengths, etc., and so
the R code for renaming became a mess of special cases. After a while, I just
gave up on decoding the excel file with R code, instead creating csv files
semi-manually, by copy-pasting into this file to do some calculations, then
copy-pasting back to csv.  I outline some of the steps here but, as is always
the case for non-automatic work, this record is best regarded as a rough
sketch.

The names were altered in an attempt to improve consistency within these
datasets and also with the Dewey datasets.  In this second category, for
example, I removed double spaces, and replaced `"` characters with the string
`in`, for inches. The original names are stored in the column named
`originalNames` and are verbatim from the excel file, *except* that multiple
consecutive spaces are converted to single spaces, leading and trailing spaces
are removed, and trailing asterisks (for footnotes) are removed.

# Details

## `chains_bio.csv` and `connectors_bio.csv`

These are green-background in the excel file, rows 7 through 12. I put `5/8"
galvanized chain` (row 11) into `chains_bio.csv` and the rest into
`connectors_bio.csv`.

The `swivel` has no listed dimensions, so I made up a height of 0.15m and a
width of 0.03m, to avoid zero height and area.

### `buoyancy`

This is inferred as `NEWTONS` (col E) divided by 9.8 m/s^2 to get kg.
```{r}
N <- c(9.3408,1.46784,0.3336,0.40032,42.3672,445)
round(N/9.8, 3)
```

### `height`

The is copied directly from col B, with 0 for the first 4 entries.

### `width`

This is inferred as the ratio of `A (m*2/m)` (col G) to `LENGTH` (col C)
```{r}
G <- c(NA,NA,NA,NA,0.05715,0.0254)
C <- c(NA,NA,NA,NA,1,1)
length <- round(G/C, 4)
length
```

### `CD`

The `CD` is assigned the value 1.3, matching the Dewey assumption for chain elements. 

### Results

These are in the order of the spreadsheet; note that `chains("?")` merges all
datasets, and orders by name.

```{r}
knitr::kable(read.csv("chains_bio.csv"), caption="Contents of chains_bio.csv")
```

and here are the connectors:

```{r}
knitr::kable(read.csv("connectors_bio.csv"), caption="Contents of connectors_bio.csv")
```



## `floats_bio.csv`

These are yellow-background in the excel file.

The first item, named `"new glass streamlined float"`, has no listed weight in
pounds or newtons, but a personal communication suggests using buoyancy 100
pounds, i.e. 445N i.e. 45.41kg.

For buoyancy, I took `NEWTONS` (col E) and divided by 9.8m/s^2 to get kg, rounding
to 2 digits after the decimal place.
```{r}
N <- c(445, 445,285,560,516,429,773,414,1886,3963,4030,2095,8224,7740)
round(N/9.8, 2)
```

There are two length columns in the file, but the values are either identical
or relatable by rounding, so I could choose either.  But is that in the x or
the y direction?

Two columns contain things related to area, colG labelled `A (m*2/m)` and colI
labelled `AW (m*2)`.  From
https://github.com/dankelley/mooring/issues/24#issuecomment-857715929, we know
that the frontal area is given by the product of colC and colG.

We take as a test case: the `"BUB 2x178" GLASS"` value, in row 42.  The
above-stated URL indicates that a BUB is a package with two 17in glass spheres,
with height (length) 1.187m (which matches colC) and frontal area 0.5531 m^2.
As a check, the product of colC (1.187) and colG (0.466) is `r 0.466*1.187`m^2,
which matches the value stated on the above-stated URL.

So, the procedure is (a) compute are as colC times colG, (b) compute diameter
as as `2*sqrt(area/pi)`.

```{r}
# Notice that row 37 is ignored, because no weight is listed for it
colC <- c(0.87, 0.74,0.563,1,0.563,1.187,1,1,1.286,1.286,1.286,1.286,2.42,2.42)
colG <- c(0.21, 0.22,0.22,0.22,0.22,0.466,0.44,0.65,1.822,1.822,1.807,1.822,0.51,0.51)
area <- colG*colC
round(area, 3)
diameter <- 2 * sqrt(colC * colG / pi)
round(diameter, 3)
```

```{r}
f <- read.csv("floats_bio.csv")
knitr::kable(f, caption="Contents of floats_bio.csv")
```

## `instruments_bio.csv`

### Processing

The entry for `DVS CURRENT METER` had height listed as `0.7 (2)`. The
parenthetic value is not explained, and so it is ignored here, perhaps
incurring an error.

The entry for `SBE37 MICROCAT CLAMP-ON STYLE*` (row 33) lists the value `(15)*`
for Newtons, and the value 15N is used here, based on my interpretation of the
footnote.

All names are converted to lower-case, and double spaces are converted to
single spaces.

The entries in rows 34 and 35 seem to refer to one item. I ignore row 34, since
so many things are not filled in for it, and I do not understand what the
meaning is.

The `height` value is copied from the first `LENGTH` column (col C).

The `buoyancy` value is inferred by dividing the `NEWTONS` (col E) by 9.8m/s^2.
```{r}
E <- c(73.392,106.752,131.216,177.92,44.48,200.16,444.8,209.056,199,192,21.8,45.6,600.48,15,144,216)
round(E/9.8, 2)
```

The `area` is computed as the product of `LENGTH` (col C) and `A (m&2/m)` (col G).

The `CD` is assigned the value 0.65, matching the Dewey assumption for floats.

A `code` entry is made, to make this compatible with Dewey values, if I can
find them.

### Results

```{r}
knitr::kable(read.csv("instruments_bio.csv"), caption="Contents of instruments_bio.csv")
```


## `releases_bio.csv`

Note the omission of an entry for type `BENTHOS 875 RELEASE`, because it has no
value listed in col G (area `A (m*2/m)`) or col I (area `AW (m*2)`).

### `height`
Use `LENGTH` (col C).

### `buoyancy`
Use the negative of `NEWTONS` (col E), divided by g=9.8m/s^2, rounded to 2 decimal
places.
```{r}
N <- c(149, 253.5, 360,	129, 273, 118.7)
round(N/9.8, 2)
```

### `width`
Use the ratio of `A (m&2/m)` (col G) to `LENGTH` (col C).
```{r}
G <- c(0.14,0.14,0.127,0.112,0.199,0.237)
C <- c(1.194,1.32,1.23,0.68,1,0.4)
width <- round(G/C, 4)
```
Note the `NA` value -- this is "benthos 875 release", which we will not use
here, since there's no point in guessing.  (Put another way, if someone at BIO
wants to take a measuring tape to that thing to find it's diameter, we'll
incorporate it!)

### `CD`
Use 0.65, i.e. Dewey's value for floats, on the assumption that the two might
be similar.

### Result for releases

```{r}
knitr::kable(read.csv("releases_bio.csv"), caption="Contents of releases_bio.csv")
```

## wires_bio.csv


### `buoyancyPerMeter`

The 3/16 case has -0.76 in a column named "W Nt/m", so I've converted that to
-0.76/9.8=-0.0776 (rounded to 4 places), to get kg/m instead of N/m.

There is no listing for the 1/4 case, but I estimated (guessed) that as

```{r}
round(-0.0776 * ((4/16)/(3/16))^2, 4)
```
This is based on an assumption that the metal inside is similar, etc.

### `diameter`

For the diameter, I looked at the column labeled `A (m*2/m)`. I assume that is
the horizontally-projected area. To test that, I computed 
```{r}
16*c(0.00635,0.0079375)/0.0254
```
which yields the expected, for the outside diameters in 16-ths of inch.

### `CD`

There are no entries in the file and I did not know what `DRAG -N` and `DRAG
-T` meant and so I just used the value $C_D=1.3$, used in Dewey's work.

### Result for `wires`

```{r}
knitr::kable(read.csv("wires_bio.csv"), caption="Contents of wires_bio.csv")
```


